# Drug data in the OMOP CDM

The OMOP CDM is a domain organised Common Data Model. Clinical data (events) are recorded into different Clinical tables depending on its domain: for example, drugs records (Drug domain) are recorded in the Drug Exposure table or condition records (Condition domain). In the following table you can see 

```{r}
cdm <- omock::mockCdmFromDataset(datasetName = "synthea-covid19-10k", source = "duckdb")
ingredients <- cdm$concept |>
  dplyr::filter(concept_class_id == "Ingredient")
drugs <- cdm$concept |>
  dplyr::filter(domain_id == "Drug", standard_concept == "S")
```

number of records per vocabulary

```{r}
x <- drugs |>
  dplyr::group_by(vocabulary_id) |>
  dplyr::tally() |>
  dplyr::collect()
ggplot2::ggplot(data = x, mapping = ggplot2::aes(x = vocabulary_id, y = n, fill = vocabulary_id)) +
  ggplot2::geom_col() +
  ggplot2::theme(legend.position = "top")
```

class concept id

```{r}
x <- drugs |>
  dplyr::group_by(concept_class_id) |>
  dplyr::tally() |>
  dplyr::collect()
ggplot2::ggplot(data = x, mapping = ggplot2::aes(x = concept_class_id, y = n, fill = concept_class_id)) +
  ggplot2::geom_col() +
  ggplot2::theme(legend.position = "top")
```

number of ingredients per concept

```{r}
x <- drugs |>
  dplyr::inner_join(
    cdm$concept_ancestor |>
      dplyr::inner_join(
        ingredients |>
          dplyr::select("ancestor_concept_id" = "concept_id"),
        by = "ancestor_concept_id"
      ) |>
      dplyr::select("concept_id" = "descendant_concept_id", "ancestor_concept_id"),
    by = "concept_id"
  ) |>
  dplyr::group_by(concept_id) |>
  dplyr::summarise(n = dplyr::n_distinct(ancestor_concept_id)) |>
  dplyr::right_join(drugs, by = "concept_id") |>
  dplyr::mutate(number_ingredients = dplyr::coalesce(n, 0)) |>
  dplyr::mutate(number_ingredients = dplyr::if_else(
    number_ingredients < 10, as.character(as.integer(number_ingredients)), "10+"
  )) |>
  dplyr::group_by(number_ingredients) |>
  dplyr::tally() |>
  dplyr::collect() |>
  dplyr::mutate(
    number_ingredients = factor(
      number_ingredients, c("0", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10+")
    ),
    percentage = sprintf("%.1f", 100 * n / sum(n)),
    percentage = dplyr::if_else(percentage == "0.0", "<0.1%", paste0(percentage, "%"))
  )
ggplot2::ggplot(data = x, mapping = ggplot2::aes(x = number_ingredients, y = n, fill = number_ingredients, label = percentage)) +
  ggplot2::geom_col() +
  ggplot2::theme(legend.position = "top") +
  ggplot2::geom_text()
```

number of brands per concept
```{r}
x <- drugs |>
  dplyr::inner_join(
     cdm$concept_relationship |> 
       dplyr::filter(relationship_id == "Has brand name") |> 
       dplyr::rename("concept_id" = "concept_id_1"),
    by = "concept_id"
  ) |>
  dplyr::group_by(concept_id) |>
  dplyr::summarise(n = dplyr::n_distinct(concept_id_2)) |>
  dplyr::right_join(drugs, by = "concept_id") |>
  dplyr::mutate(number_brands = dplyr::coalesce(n, 0)) |>
  dplyr::mutate(number_brands = dplyr::if_else(
    number_brands < 10, as.character(as.integer(number_brands)), "10+"
  )) |>
  dplyr::group_by(number_brands) |>
  dplyr::tally() |>
  dplyr::collect() |>
  dplyr::mutate(
    number_brands = factor(
      number_brands, c("0", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10+")
    ),
    percentage = sprintf("%.1f", 100 * n / sum(n)),
    percentage = dplyr::if_else(percentage == "0.0", "<0.1%", paste0(percentage, "%"))
  )
ggplot2::ggplot(data = x, mapping = ggplot2::aes(x = number_brands, y = n, fill = number_brands, label = percentage)) +
  ggplot2::geom_col() +
  ggplot2::theme(legend.position = "top") +
  ggplot2::geom_text()
```

Explain example of COVID-19 vaccination with brands.

number of dose forms

circular bar plot

dose form is only available with RxNorm

There is a minority of concepts (6926, 0.3%) has associated 2 dose forms, most of this records come from an overlap between Injectable Solution, Ijection and Intravenous solution, but this is a minority and in general records will only have one dose (or none) associated.

```{r}
x <- drugs |>
  dplyr::inner_join(
     cdm$concept_relationship |> 
       dplyr::filter(relationship_id == "RxNorm has dose form") |> 
       dplyr::rename("concept_id" = "concept_id_1"),
    by = "concept_id"
  ) |>
  dplyr::group_by(concept_id) |>
  dplyr::summarise(n = dplyr::n_distinct(concept_id_2)) |>
  dplyr::right_join(drugs, by = "concept_id") |>
  dplyr::mutate(number_dose_form = dplyr::coalesce(n, 0)) |>
  dplyr::mutate(number_dose_form = dplyr::if_else(
    number_dose_form < 10, as.character(as.integer(number_dose_form)), "10+"
  )) |>
  dplyr::group_by(number_dose_form) |>
  dplyr::tally() |>
  dplyr::collect() |>
  dplyr::mutate(
    number_dose_form = factor(
      number_dose_form, c("0", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10+")
    ),
    percentage = sprintf("%.1f", 100 * n / sum(n)),
    percentage = dplyr::if_else(percentage == "0.0", "<0.1%", paste0(percentage, "%"))
  )

drugs |>
  dplyr::inner_join(
     cdm$concept_relationship |> 
       dplyr::filter(relationship_id == "RxNorm has dose form") |> 
       dplyr::rename("concept_id" = "concept_id_1"),
    by = "concept_id"
  ) |>
  dplyr::group_by(concept_id) |>
  dplyr::filter(dplyr::n_distinct(concept_id_2) == 2) |>
  dplyr::select("concept_id", "concept_id_2") |>
  PatientProfiles::addConceptName(column = "concept_id_2", nameStyle = "route") |>
  dplyr::select(!"concept_id_2") |>
  dplyr::collect() |>
  dplyr::arrange(concept_id, route) |>
  dplyr::mutate(id = paste0("route_", dplyr::row_number())) |>
  tidyr::pivot_wider(names_from = "id", values_from = "route") |>
  dplyr::mutate(route = paste0(route_1, "; ", route_2)) |>
  dplyr::group_by(route) |>
  dplyr::tally()

ggplot2::ggplot(data = x, mapping = ggplot2::aes(x = number_dose_form, y = n, fill = number_dose_form, label = percentage)) +
  ggplot2::geom_col() +
  ggplot2::theme(legend.position = "top") +
  ggplot2::geom_text()
```

number of route categories

number ATC classes 
